### 2025-03-12 paper/quarto_ecml/paper.tex
Review related literature

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

To the best of our knowledge, our proposed framework of counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow \cite{augustin2020adversarial} in tying the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e., adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

@@

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate the augmenting the training data of
image classifiers can improve generalization. \cite{teney2020learning}
propose an approach using counterfactuals in training that does not rely
on data augmentation: they argue that counterfactual pairs typically
already exist in training datasets. Specifically, their approach relies
on identifying similar input samples with different annotations and
ensuring that the gradient of the classifier aligns with the vector
between such pairs of counterfactual inputs using the cosine distance as
the loss function.

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation:
\cite{wu2021polyjuice}, propose \emph{Polyjuice}, a general-purpose
counterfactual generator for language models. They demonstrate
empirically that augmenting training data through \emph{Polyjuice}
counterfactuals improves robustness in a number of NLP tasks.
\cite{balashankar2023improving} also use \emph{Polyjuice} to augment NLP
datasets through diverse counterfactuals and show that classifier
robustness improves up to 20\%. Finally, \cite{luu2023counterfactual}
introduce Counterfactual Adversarial Training (CAT), which also aims at
improving generalization and robustness of language models.
Specifically, they propose to proceed as follows: firstly, they identify
training samples that are subject to high predictive uncertainty;
secondly, they generate counterfactual explanations for those samples;
and, finally, they fine-tune the given language model on the augmented
dataset that includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problem in representation learning. Formally,
AE can be seen as the subset of CE for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CE and AE are equivalent under
certain conditions and derive theoretical upper bounds on the distances
between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that are
guaranteed to provide recourse for individuals to move from an adverse
outcome to some positive target class with high probability. Their
approach builds on adversarial training, where in this context
susceptibility to targeted adversarial examples for the positive class
is explicitly induced. The proposed method allows for imposing a set of
actionability constraints ex-ante: for example, users can specify that
certain features (e.g., \emph{age}, \emph{gender}, \ldots) are
immutable. Secondly, \cite{guo2023counternet} are the first to propose
an end-to-end training pipeline that includes counterfactual
explanations as part of the training procedure. In particular, they
propose a specific network architecture that includes a predictor and CE
generator network, where the parameters of the CE generator network are
learnable. Counterfactuals are generated during each training iteration
and fed back to the predictor network. In contrast to
\cite{guo2023counternet}, we impose no restrictions on the neural
network architecture at all.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally \textbf{faithful} to the model in that they are consistent
with what the model has learned about the underlying data. Formally,
they rely on tools from energy-based modelling to minimize the
divergence between the distribution of counterfactuals and the
conditional posterior over inputs learned by the model. Their proposed
counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
and only if the underlying model has learned representations that align
with them. They find that both deep ensembles
\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\cite{grathwohl2020your} tend to do well in this regard.

>>>>

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

To the best of our knowledge, the proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements; the former means that
the counterfactuals should comply with the distribution of the factual
data, the latter means that they should respect arbitrary (global)
actionability constraints. To make the desiderata for our framework more
concrete, we follow \cite{augustin2020adversarial} in tying the concept
of explainability to the quality of counterfactual explanations that we
can generate for a given model. The authors show that CEs---understood
here as minimal input perturbations that yield some desired model
prediction---are generally more meaningful if the underlying model is
more robust to adversarial examples. We can make intuitive sense of this
finding when looking at adversarial training (AT) through the lens of
representation learning with high degrees of freedom. As argued before,
learned representations may be sensitive to producing implausible
explanations and mispredicting for worst-case counterfactuals (i.e.,
adversarial examples). Thus, by inducing models to ``unlearn''
susceptiblity to such examples, AT can effectively remove implausible
explanations from the solution space.

@@

This interpretation of the link between explainability through
counterfactuals on one side and robustness to adversarial examples on
the other is backed by empirical evidence.
\cite{sauer2021counterfactual} demonstrate that using counterfactual
images during classifier training improves model robustness. Similarly,
\cite{abbasnejad2020counterfactual} argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate that augmenting the training data of
image classifiers can improve generalization. Finally,
\cite{teney2020learning} propose an approach using counterfactuals in
training that does not rely on data augmentation: they argue that
counterfactual pairs typically already exist in training datasets.
Specifically, their approach relies on identifying similar input samples
with different annotations and ensuring that the gradient of the
classifier aligns with the vector between such pairs of counterfactual
inputs using the cosine distance as the loss function.

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation.
\cite{wu2021polyjuice} propose \emph{Polyjuice}, a general-purpose
counterfactual generator for language models. They demonstrate
empirically that the augmentation of training data through
\emph{Polyjuice} counterfactuals improves robustness in a number of NLP
tasks. \cite{balashankar2023improving} similarly use \emph{Polyjuice} to
augment NLP datasets through diverse counterfactuals and show that
classifier robustness improves by up to 20\%. Finally,
\cite{luu2023counterfactual} introduce Counterfactual Adversarial
Training (CAT), which also aims at improving generalization and
robustness of language models through a three-step procedure. First, the
authors identify training samples that are subject to high predictive
uncertainty. Second, they generate counterfactual explanations for those
samples. Finally, they fine-tune the given language model on the
augmented dataset that includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples (AE).
Pointing to clear similarities in how CEs and AEs are generated,
\cite{freiesleben2022intriguing} makes the case for jointly studying the
opaqueness and robustness problems in representation learning. Formally,
AEs can be seen as the subset of CEs for which misclassification is
achieved \cite{freiesleben2022intriguing}. Similarly,
\cite{pawelczyk2022exploring} show that CEs and AEs are equivalent under
certain conditions and derive theoretical upper bounds on distances
between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of the post-hoc counterfactual explanations. Firstly,
\cite{ross2021learning} propose a way to train models that guarantee
individual recourse to some positive target class with high probability.
Their approach builds on adversarial training by explicitly inducing
susceptibility to targeted adversarial examples for the positive class.
Additionally, the proposed method allows for imposing a set of
actionability constraints ex-ante. For example, users can specify that
certain features (e.g., \emph{age}, \emph{gender}) are immutable.
Secondly, \cite{guo2023counternet} are the first to propose an
end-to-end training pipeline that includes counterfactual explanations
as part of the training procedure. In particular, they propose a
specific network architecture that includes a predictor and CE generator
network, where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network. In contrast to \cite{guo2023counternet},
we impose no restrictions on the neural network architecture at all.

@@

Improving the adversarial robustness of models is not the only path
towards aligning representations with plausible explanations. In a work
closely related to this one, \cite{altmeyer2024faithful} show that
explainability can be improved through model averaging and refined model
objectives. The authors propose a way to generate counterfactuals that
are maximally faithful to the model in that they are consistent with
what the model has learned about the underlying data. Formally, they
rely on tools from energy-based modelling to minimize the divergence
between the distribution of counterfactuals and the conditional
posterior over inputs learned by the model. Their proposed
counterfactual explainer, \emph{ECCCo}, yields plausible explanations if
and only if the underlying model has learned representations that align
with them. The authors find that both deep ensembles
\cite{lakshminarayanan2016simple} and joint energy-based models (JEMs)
\cite{grathwohl2020your} tend to do well in this regard.

### 2025-03-12 paper/sections/related.qmd
Review related literature

To the best of our knowledge, our proposed framework of counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability. In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements. To make this more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of counterfactual explanations that we can generate for a given model. The authors show that counterfactual explanations---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom: by inducing models to "unlearn" representations that are susceptible to worst-case counterfactuals (i.e., adversarial examples), AT effectively removes some implausible explanations from the solution space.

@@

This interpretation of the link between explainability through counterfactuals on one side, and robustness to adversarial examples on the other, is backed by empirical evidence. @sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, demonstrate the augmenting the training data of image classifiers can improve generalization. @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets. Specifically, their approach relies on identifying similar input samples with different annotations and ensuring that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using the cosine distance as the loss function.

In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose *Polyjuice*, a general-purpose counterfactual generator for language models. They demonstrate empirically that augmenting training data through *Polyjuice* counterfactuals improves robustness in a number of NLP tasks. @balashankar2023improving also use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves up to 20%. Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models. Specifically, they propose to proceed as follows: firstly, they identify training samples that are subject to high predictive uncertainty; secondly, they generate counterfactual explanations for those samples; and, finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples (AE). Pointing to clear similarities in how CE and AE are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problem in representation learning. Formally, AE can be seen as the subset of CE for which misclassification is achieved [@freiesleben2022intriguing]. Similarly, @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive theoretical upper bounds on the distances between them.

Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of post-hoc counterfactual explanations. Firstly, @ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals to move from an adverse outcome to some positive target class with high probability. Their approach builds on adversarial training, where in this context susceptibility to targeted adversarial examples for the positive class is explicitly induced. The proposed method allows for imposing a set of actionability constraints ex-ante: for example, users can specify that certain features (e.g., *age*, *gender*, ...) are immutable. Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes counterfactual explanations as part of the training procedure. In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network. In contrast to @guo2023counternet, we impose no restrictions on the neural network architecture at all.

@@

Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations. In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives. The authors propose a way to generate counterfactuals that are maximally **faithful** to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. They find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.

>>>>

To the best of our knowledge, the proposed framework for counterfactual training represents the first attempt to use counterfactual explanations during training to improve model explainability. In high-level terms, we define model explainability as the extent to which valid explanations derived for an opaque model are also deemed plausible with respect to the underlying data and stakeholder requirements; the former means that the counterfactuals should comply with the distribution of the factual data, the latter means that they should respect arbitrary (global) actionability constraints. To make the desiderata for our framework more concrete, we follow @augustin2020adversarial in tying the concept of explainability to the quality of counterfactual explanations that we can generate for a given model. The authors show that CEs---understood here as minimal input perturbations that yield some desired model prediction---are generally more meaningful if the underlying model is more robust to adversarial examples. We can make intuitive sense of this finding when looking at adversarial training (AT) through the lens of representation learning with high degrees of freedom. As argued before, learned representations may be sensitive to producing implausible explanations and mispredicting for worst-case counterfactuals (i.e., adversarial examples). Thus, by inducing models to "unlearn" susceptiblity to such examples, AT can effectively remove implausible explanations from the solution space.

@@

This interpretation of the link between explainability through counterfactuals on one side and robustness to adversarial examples on the other is backed by empirical evidence. @sauer2021counterfactual demonstrate that using counterfactual images during classifier training improves model robustness. Similarly, @abbasnejad2020counterfactual argue that counterfactuals represent potentially useful training data in machine learning, especially in supervised settings where inputs may be reasonably mapped to multiple outputs. They, too, demonstrate that augmenting the training data of image classifiers can improve generalization. Finally, @teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets. Specifically, their approach relies on identifying similar input samples with different annotations and ensuring that the gradient of the classifier aligns with the vector between such pairs of counterfactual inputs using the cosine distance as the loss function.

In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation. @wu2021polyjuice propose *Polyjuice*, a general-purpose counterfactual generator for language models. They demonstrate empirically that the augmentation of training data through *Polyjuice* counterfactuals improves robustness in a number of NLP tasks. @balashankar2023improving similarly use *Polyjuice* to augment NLP datasets through diverse counterfactuals and show that classifier robustness improves by up to 20%. Finally, @luu2023counterfactual introduce Counterfactual Adversarial Training (CAT), which also aims at improving generalization and robustness of language models through a three-step procedure. First, the authors identify training samples that are subject to high predictive uncertainty. Second, they generate counterfactual explanations for those samples. Finally, they fine-tune the given language model on the augmented dataset that includes the generated counterfactuals.

There have also been several attempts at formalizing the relationship between counterfactual explanations and adversarial examples (AE). Pointing to clear similarities in how CEs and AEs are generated, @freiesleben2022intriguing makes the case for jointly studying the opaqueness and robustness problems in representation learning. Formally, AEs can be seen as the subset of CEs for which misclassification is achieved [@freiesleben2022intriguing]. Similarly, @pawelczyk2022exploring show that CEs and AEs are equivalent under certain conditions and derive theoretical upper bounds on distances between them.

Two recent works are closely related to ours in that they use counterfactuals during training with the explicit goal of affecting certain properties of the post-hoc counterfactual explanations. Firstly, @ross2021learning propose a way to train models that guarantee individual recourse to some positive target class with high probability. Their approach builds on adversarial training by explicitly inducing susceptibility to targeted adversarial examples for the positive class. Additionally, the proposed method allows for imposing a set of actionability constraints ex-ante. For example, users can specify that certain features (e.g., *age*, *gender*) are immutable. Secondly, @guo2023counternet are the first to propose an end-to-end training pipeline that includes counterfactual explanations as part of the training procedure. In particular, they propose a specific network architecture that includes a predictor and CE generator network, where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network. In contrast to @guo2023counternet, we impose no restrictions on the neural network architecture at all.

@@

Improving the adversarial robustness of models is not the only path towards aligning representations with plausible explanations. In a work closely related to this one, @altmeyer2024faithful show that explainability can be improved through model averaging and refined model objectives. The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their proposed counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your] tend to do well in this regard.
